---
title: "Flights of New York"
author: Gowri Prathap
documentclass: article
geometry: margin=1in
fontsize: 11pt
output:
  pdf_document:
    toc: false
    df_print: tibble
    fig_caption: false
    number_sections: false
    dev: pdf
    highlight: tango
  html_document:
    theme: default
    self_contained: true
    toc: false
    df_print: tibble
    fig_caption: false
    number_sections: false
    smart: true
    dev: svg
---

```{r setup, include = FALSE}
# DO NOT ALTER THIS CHUNK
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  fig.width = 5,
  fig.asp = 0.618,
  out.width = "70%",
  dpi = 120,
  fig.align = "center",
  cache = FALSE
)
# Load required packages
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(nycflights13))
```

Type the following command in the Console window to view the dataset:

View(flights)

# Important questions about te dataset: 

1. How many rows and columns does this dataset have? 
This data set has are 336776 rows and 19 columns.

2. What does a single row in this dataset represent? 
Each row in this dataset represents data for each flight that departed NYC (i.e. JFK, LGA or EWR) in 2013.

3. What is the difference between the information contained in the dep_time and sched_dep_time columns? 
The dep_time gives the actual departing time of the flights, while sched_dep_time gives the scheduled departing time of the flights.

4. Which columns contain information about dates and times? 
Column 3 (day) gives the date and columns 15-19 give the air times in hours, minutes and so on. Columns 4-9 give the timings of the departure, arrival, and so on.

5. Airplanes are reused across many different flights. Which columns would be helpful to use in identifying individual airplanes? 
The 'flight' column and 'tailnum' column which gives the flight number and plane tail number respectively will help identify individual flights.

# Important functions:

## select() function 

The select() function selects columns from a dataset.
```{r}
flights %>%
  select(year, month)
```
This command displays the columns year and month since it was included in the select() function.
The symbol %>% is called the pipe and is used to pass a dataset through a chain of commands.

```{r}
flights %>%
  select(year:day)
```
The colon prints all columns between including the two columns mentioned with the colon. Here it prints all the columns form year to month since the command says year:day.


## arrange() function

The arrange() function sorts columns with textual data (chr data type) into alphabetical order and sorts numerical data into numerical order.

```{r}
flights %>%
  arrange(month, day)
```

```{r}
flights %>%
  arrange(day, month)
```
Based on the output, it looks like both the month and day columns were sorted. Day column was sorted first.
If we reverse the order of columns in arrange(), the first piece of code does not make a difference, since days were arranged first, and days always change before months so it does not differ from the original dataset. The second piece of code makes a difference if reversed because it sorts the months first, in the order 1, 2, 3,...etc. and days only after that. Hence it is different from the first dataset.

The column that gives the departure delay is 'dep_delay'.
```{r}
flights %>%
  arrange(desc(dep_delay))
```
The first row of the arranged dataset has the longest departure delay, which is flight 51 tail number N384HA.


## mutate() function

mutate() lets us transform a dataset by applying the same operation to each row in the dataset and appending the results as a new column.
```{r}
flights %>%
  mutate(
    average_speed = distance / (air_time * 60)
  )
```
The new column shows up as the last column of the dataset. The name of this new column is average_speed. The part of the code under mutate() where the column is declared, where the column name is given, is taken as the column name. Here, under mutate() it is given as average_speed = the formula, and hence average_speed is taken as the name of the column.

```{r}
flights %>%
  mutate(
    dep_time_hour = dep_time %/% 100,
    dep_time_minute = dep_time %% 100,
    dep_time_minutes_midnight = dep_time_hour * 60 + dep_time_minute
  )
```
We used the  dep_time_hour and dep_time_minute columns to compute the number of minutes since midnight.

## filter() function

To find all the flights operated by United Airlines (UA) that arrived early.
```{r}
flights %>%
  filter(
    arr_delay < 0
  )
```

```{r}
flights %>%
  filter(
    carrier == "UA"
  )
```

This code can be combined into a single block of code:
```{r}
flights %>%
  filter(
    arr_delay < 0,
    carrier == "UA"
  )
```

##Using group_by() and summarize()

To calculate the mean (average) arrival delay for each airline carrier:
```{r}
flights %>%
  group_by(carrier) %>%
  summarize(
    average_arr_delay = mean(arr_delay, na.rm = TRUE)
  )
```
Airline carrier FL had the longest arrival delays on average. The airline carrier AS had the shortest arrival delays on average.

